Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-23 manuscript.Rmd
make edu case more prominent in the intro
Background III

The shortcoming of current debates is that they focus on whether people *ever* change, a question that is not theoretically interesting since the answer is almost always “yes.”  As noted previously, if researchers observe enough people for long enough, they will likely observe some of them making durable changes. A lack of evidence for change in one question might simply indicate that the question was not asked at a clear enough level of resolution, for a long enough time frame, or to enough people to find evidence of change. Conversely, finding evidence of change tells us little about how much change is present in a population and how meaningful that change is for social differences.

A more theoretically productive approach is to start with a model that assumes that, in the window of a panel survey, some people might remain stable, some people might change a little, and some people might undergo substantial changes in personal culture. Lersch’s LCAM formally does this in modeling each person as following their own trajectory over time. But rather than adjudicating whether this model fits the data better than one in which people never change –-asking the question of whether change ever exists-– researchers should quantify the amount that people change over time and compare that change to the differences that exist between people at the beginning of the panel. In other words, rather than ask, “do people change?” we should ask, “what are the relative contributions of change and initial differences toward explaining differences in personal culture?”

In this approach, there are two numbers that each reflect a combination of relevant theoretical processes. First, interpersonal differences at the start of the survey, what Lersch calls “early imprinting,” reflects the accumulated experiences of people prior to entering the panel survey that shape how they answer questions. While this is typically associated with a formative period or a settled disposition (hence the name early imprinting), it could also reflect transformative experiences that happen at any age, so long as they occurred before the start of the panel and leave a lasting effect on how people answer questions. For example, among people already retired upon entering the panel, their “early imprinting” might reflect this life course transition. To some extent, then, this number also reflects variation across people in relevant social roles or statuses at the start of the panel that are important in shaping dispositions.

The second set of processes, what Lersch calls “persistent change” or “adaption” and Kiley and Vaisey call “active updating,” reflects durable changes in personal culture over time and is captured by the amount of intrapersonal linear change over the course of the survey. Lersch suggests that these changes are brought on by social triggers such as moving into a new environment or adopting new social roles. They could also reflect the diffusion of new culture across social networks, changes in social network composition (or knowledge about who is in a person's network), new cues from political elites or other cultural leaders, the emergence of new issues in politics or culture, or other large-scale social shifts.

A third number, often called “residual variance” or “fluctuation,” necessarily accounts for the remainder of variance in peoples’ responses over time. This non-durable change potentially emerges for a variety of reasons. For example, people might not have clear personal cultural dispositions as measured by surveys. Instead, they might internalize a broad set of considerations and construct an opinion or attitude in the context of the survey interview, with different considerations coming to the forefront of their cognition each interview [@tourangeau2000; @zaller1992; @feldman1992]. This variance in considerations can produce variance in responses even if peoples’ “dispositions” --the full set of considerations they could draw on-- remains stable over time. This variance could also include true errors of measurement, such as respondents misunderstanding the question, respondents or administrators selecting the wrong response, or responses getting coded incorrectly. While this third process reflects important aspects of personal culture, it does not directly touch on the ongoing debate outlined here. Because of this we quantify this component but focus principally on the other two.

We assert that, at the theoretical level, the existence of these two principal processes --interpersonal differences and intrapersonal change-- is indisputable, and that they are, in many ways, related. If intrapersonal change occurs at any point in the life course, it will inevitably produce interpersonal differences by the time people enter a panel survey. Similarly, unless people are fully socialized early in the life course and never deviate from these dispositions (a proposition against which there is substantial evidence), then we expect to observe intrapersonal change in some proportion of the population whenever we survey them over time.

Furthermore, these numbers are not necessarily useful in isolation or without a sense of time. What is important for the theoretical debate is the relative contribution of these two components in explaining cultural differences in a population. In other words, when we look at a population of adults over some time frame, is there enough intrapersonal change relative to interpersonal difference to view it as an important explanation for why people demonstrate cultural differences? That is a question that can only be answered by quantifying the two processes.

However, the real value in quantification lies not in adjudicating past debates, but in opening up avenues of exploration that were limited under previous approaches. For example, classifying questions as either displaying durable change or not obscures differences among questions where there is evidence of durable change, but the relative importance of durable change is different. In quantifying the relative contributions of interpersonal differences and intrapersonal change, researchers can compare the relative importance of these processes in explaining differences across societies, across groups within the same population, across items in the same survey, and across time.

>>>>

Current debates often narrow down to ask whether people *ever* change, which is rarely theoretically interesting since the answer is almost always “yes.”  Over sufficient time and in a large enough sample, researchers will likely observe some evidence for durable changes in individuals. A lack thereof may simply due to a lack of clarity in the question asked or a low level of resolution in the possible answers, or that the question simply was not asked long enough or to enough people. Conversely, finding evidence of change tells us little about how much change is present in a population and how meaningful that change is for social differences.

A more theoretically productive approach begins with a model assuming that during an observed time period some individuals might remain stable, might change a little, or might undergo significant shifts in their personal culture. Lersch’s LCAM formally does this by modeling each individual as following their own trajectory over time. But rather than debating whether this model is superior to a model that assumes that people never change, researchers should quantify individual changes over time and compare those to the differences that exist between people at the panel's start. In other words, the focus should shift from asking, “do people change?” to asking, “how do change and initial differences collectively explain variation in personal culture?” [I CHANGED THAT QUESTION WORDING, PLEASE CHECK!]

In this approach, two metrics reflect distinct theoretical processes. First, "early imprinting," as Lersch calls it, denotes interpersonal differences at the start of the panel, capturing the accumulated experiences of people before participating in the panel that influence their answers. While commonly associated with experiences during an early formative period that result in settled disposition, this could also reflect transformative experiences [-@paul2014transformative] that happen at any stage in life, as long as they predate the panel and consistently impact subsequent responses. For instance, for those people entering the panel post-retirement, this “imprinting” might reflect this pivotal life transition. Consequently, this metric also reflects variation in individuals' social roles or statuses at the start of the panel that were important in shaping their dispositions.

The second process, termed “persistent change” or “adaption” by Lersch and "active updating" by Kiley and Vaisey, denotes durable changes in personal culture over time, captured by the amount of intrapersonal linear change during the panel. Lersch attributes these changes to social triggers such as moving into a new environment or adopting new social roles. Additionally, they might reflect the diffusion of new cultural forms across social networks, cues from political elites or otherwise culturally influential leaders, the emergence of issues in politics or culture, or large-scale social shifts.

A third metric, often termed “residual variance” or “fluctuation,” represents the remainder of variance in peoples’ responses over time. This non-durable change may arise for various reasons. Some might lack clear cultural dispositions, leading them to construct an opinion or attitude during the survey, with varying thoughts and considerations dominating their thinking in each interview [@tourangeau2000; @zaller1992; @feldman1992]. Even though the full set of considerations they could draw on remains stable over time, this could yield to such variance in responses. Additionally, this variance can also include measurement errors, such as misinterpreted questions, erroneous response selections, or responses getting coded incorrectly. While this third process thus might reflect potentially important aspects of personal culture [I FELT WE NEED TO PLAY IT DOWN A BIT], it is less central to the debate outlined here. Hence, we focus on the first two metrics.

At a theoretical level, the coexistence of two main processes -- interpersonal differences and intrapersonal change -- is undeniable, and they are linked in many ways [THOUGHT IF intrinsically linked]. Any intrapersonal change during one's life will inevitably manifest as interpersonal differences by the time people enter a panel survey. Furthermore, unless people are entirely socialized early on and never deviate from these dispositions -- an understanding challenged by evidence -- we expect observing intrapersonal change in a segment of the population when surveyed over time.

Furthermore, these metrics are not necessarily useful in isolation or without context. Central to the theoretical debate is how the two processes they capture collectively explain cultural differences in a population. In other words, when examining adults across a period of time, does intrapersonal change, compared to interpersonal differences, sufficiently account for the observed cultural differences? Only through quantifying the two processes we can get closer to a true answer to this question.

Yet, the true merit of the proposed quantification goes beyond settling past debates; it opens up new avenues for investigations that remained closed under previous approaches. For example, merely classifying questions based on whether they show durable change [DO WE NEED "or not"? NEVER KNOW THAT AS A NON-NATIVE] overlooks possibly important differences among those that do. By quantifying the contributions of interpersonal differences and intrapersonal change, researchers can gauge their relative importance in understanding [INSTEAD explaining] differences among survey items, between groups within a given population, across time, and even across societies.
